An optogenetic cell therapy to restore control of target muscles in an aggressive mouse model of Amyotrophic Lateral Sclerosis

Version 1

Abstract

Abstract

Breakdown of neuromuscular junctions (NMJs) is an early pathological hallmark of amyotrophic lateral sclerosis (ALS) that blocks neuromuscular transmission, leading to muscle weakness, paralysis and, ultimately, premature death. Currently, no therapies exist that can prevent progressive motor neuron degeneration, muscle denervation or paralysis in ALS. Here, we report important advances in the development of an optogenetic, neural replacement strategy that can effectively restore innervation of severely affected skeletal muscles in the aggressive SOD1G93A mouse model of ALS, thus providing an interface to selectively control the function of targeted muscles using optical stimulation. We also identify a specific approach to confer complete survival of allogeneic replacement motor neurons. Furthermore, we demonstrate that an optical stimulation training paradigm can prevent atrophy of reinnervated muscle fibers and results in a 10-fold increase in optically-evoked contractile force. Together, these advances pave the way for an assistive therapy that could benefit all ALS patients.

Introduction

The progressive degeneration of motor neurons that occurs in amyotrophic lateral sclerosis (ALS), the most common adult motor neuron disease (MND), affects almost all cellular components of the neuromuscular system, including cortical and spinal motor neurons, interneurons Crabé et al. (2020), glial cells Van Harten et al. (2021); Vahsen et al. (2021), as well as muscle Loeffer et al. (2016). Motor axons, contained within peripheral nerves, serve as the final common relay for transmission of motor signals that control all voluntary muscle contraction and movement. However, one of the earliest characteristic pathological features of ALS involves “die-back” of motor axon terminals Fischer et al. (2004) and breakdown of neuromuscular junctions (NMJs), the specialized synapses between motor axon terminals and muscle fibres. Effectively, this represents a single point of failure that permanently blocks motor signal transmission, irrespective of the condition of central motor circuits involved in coordination of movement signals. This results in an inexorable progression of muscle weakness, atrophy and, ultimately, complete paralysis, eventually leading to premature death. The median survival time in ALS, from initial onset of symptoms to death, typically as a result of respiratory complications, is only 20-48 months Chiò et al. (2009).

To date, over 20 monogenic mutations that cause ALS have been identified, although these still only account for 10% of total cases, with the remaining 90% being classified as sporadic ALS Mathis et al. (2019). The downstream cellular and molecular pathomechanisms underlying neurodegeneration in ALS are extremely complex and include dysregulation of proteostasis, autophagy, RNA metabolism and axon transport, as well as excitotoxicity, oxidative stress and neuroinflammation Mejzini et al. (2019). Given the disparate causes and complex disease mechanisms, development of an effective therapy has proven extremely challenging and there are currently no effective treatments capable of arresting the progressive paralysis that occurs in ALS Brown and Al-Chalabi (2017). Even emergent gene therapy approaches, such as antisense oligonucleotides (ASOs) that have shown early promise in clinical trials Miller et al. (2013), are unlikely to benefit the majority of patients with sporadic ALS and will not replace degenerated motor neurons or restore motor function once it has been lost.

Similarly, efforts to develop cell therapy approaches for ALS have, so far, primarily been aimed at slowing the degeneration of motor neurons in the spinal cord by providing neurotrophic factor (NTF) support through intraspinal engraftment of foetally-derived neural stem cells (NSCs) Feldman et al. (2014); Goutman et al. (2018) or intrathecal delivery of autologous mesenchymal stem cells (MSCs) that are modified to overexpress NTFs Cudkowicz et al. (2022); Berry et al. (2019). Whilst these approaches were shown to significantly slow disease progression in transgenic animal models of ALS Xu et al. (2006); Yan et al. (2006), clinical trials in ALS patients have shown only a modest therapeutic effect in the case of intraspinal NSC grafts Glass et al. (2012) and no overall benefit of intrathecal MSC delivery Cudkowicz et al. (2022) Given the highly invasive surgical laminectomy required to engraft NSCs into the ventral horn of the spinal cord, only localized populations of motor neurons in the lumbar and/or cervical region were targeted, however, the surgical procedure was generally well tolerated and the approach was proven to be safe in clinical trials Feldman et al. (2014); Glass et al. (2016). Moreover, these clinical trials have provided evidence that specific muscle functions can be preserved for longer in some ALS patients Mazzini et al. (2019). Importantly, this pioneering approach provides a precedent for implementation of an allogeneic stem cell-based therapy and also shows that ALS patients can tolerate a 6 month period of immuno-suppression Mazzini et al. (2015), which appears to be sufficient to confer long term survival of engrafted cells Tadesse et al. (2014). Nonetheless, it appears unlikely that this approach will be able to restore motor function once it has been lost, since the NSCs do not replace lost motor neurons, and any therapeutic effect has so far been shown to be transient Goutman et al. (2018); Mazzini et al. (2019). Therefore, there remains an urgent unmet need to develop novel therapies that can rescue muscle innervation and maintain muscle function in ALS patients.

We have previously demonstrated a novel proof-of-concept strategy to overcome muscle denervation and restore control of muscle contraction in a nerve injury model of muscle paralysis that could have major therapeutic value for restoring function of any targeted muscle or group of muscles in ALS patients Bryson et al. (2014). Briefly, optogenetically-modified replacement motor neurons, derived from murine embryonic stem cells (mESCs), were engrafted into distal branches of peripheral nerves supplying specific lower hindlimb flexor and extensor target muscles in wild-type mice that had undergone a nerve ligation injury. Our results showed that the engrafted motor neurons were able to project axons from the graft site to the target muscles where they formed de novo NMJs. Due to the ectopic location of the engrafted motor neurons, outside the central nervous system (CNS), they do not receive endogenous motor signals, and must therefore be exogenously activated. Expression of the blue-light sensitive channelrhodopsin-2 (ChR2) protein Nagel et al. (2003) in the engrafted motor neurons conferred the ability to selectively activate these engrafted neurons and thereby control the contractile function of the target muscle using acute optical stimulation Bryson et al. (2014). The aim of this neural replacement strategy is therefore to provide a biological interface capable of rendering any target muscle receptive to control signals transmitted by optical stimulation to engrafted motor neurons Bryson et al. (2016). Importantly, we have recently developed a prototype 64-channel stimulation and recording device capable of controlling multiple independent intraneural graft sites that could be used to elicit coordinated function of large numbers of muscles, in order to restore useful motor functions Liu et al. (2022).

This novel approach to restore control of paralyzed muscles in ALS patients, using a combination of cell replacement and optical stimulation, has several key advantages over existing cell replacement and electrical stimulation strategies, including: i) the ability to engraft motor neurons peripherally, in close proximity to targeted muscles, which greatly accelerates the rate of reinnervation and reduces the period of denervation, consequently ameliorating denervation-induced muscle atrophy; ii) avoidance of engrafting replacement cells into the neurotoxic environment that exists within the CNS of ALS patients and the necessity for reinnervating axons of CNS-engrafted motor neurons to overcome the inhibitory CNS:PNS barrier in order to exit the CNS and grow the often long distances to target muscles; iii) specificity of optical stimulation to the engrafted ChR2+ motor neurons avoids painful off-target activation of sensory afferents or aberrant activation of endogenous motor axons associated with electrical nerve stimulation (ENS); and critically, iv) the ability to recruit motor units in correct physiological order using optical nerve stimulation (ONS) Llewellyn et al. (2010), avoids the problem of rapid muscle fatigue associated with ENS and incorrect, non-physiological motor unit recruitment. Furthermore, ENS-mediated control of muscle function depends on the presence of surviving motor axons and, since these are progressively lost during the course of disease progression in ALS, the ability of ENS to induce muscle contraction is steadily eroded. More importantly, it has recently been shown that ENS, applied to the phrenic nerve to assist respiratory function in two separate clinical trials in ALS patients McDermott et al. (2016); Gonzalez-Bermejo et al. (2016) accelerated diaphragm muscle denervation, which likely contributed to a significant reduction in patient life-span Guimarães-Costa et al. (2019). Therefore, it is unlikely that ENS will ever be suitable for artificial control of critical muscle function in ALS patients. In contrast, optogenetic stimulation Henderson et al. (2009), in combination with neural replacement, represents a safe alternative approach to artificially restore innervation and function of paralyzed muscles in ALS.

In the current study, we sought to optimize critical elements of this novel therapeutic strategy and to determine whether ChR2+ motor neurons can be used to successfully restore innervation, induce muscle contraction and prevent atrophy of targeted muscles in a model of ALS using the highly aggressive SOD1G93A mouse model of ALS.

Results

Host-vs-graft rejection causes loss of most intraneural ESC-MNs allografts

ChR2+ motor neurons were derived from our previously characterized mouse embryonic stem cell (mESC) line Bryson et al. (2014) and differentiated using a well-established protocol Wichterle et al. (2002). Since these donor ChR2+ motor neurons were generated from an mESC line originating from the 129S1SvImJ mouse strain (Supplementary Table 1) and recipient mice SOD1G93A mice were on a congenic C57BL/6J genetic background, they constitute an allogeneic source of donor cells. Importantly, allogeneic cells are likely to provide a more cost-effective, off-the-shelf cell therapy platform, compared to generating individual, patient-specific batches of cells suitable for autologous engraftment. Therefore, an important initial objective of this study was to identify an immunosuppression regimen capable of preventing host-vs-graft rejection of allogeneic ChR2+ motor neurons to enable their innervation of target muscles in recipient SOD1G93A mice.

In the absence of any immunosuppression, the survival rate of ChR2+ motor neuron grafts at 35 d post engraftment, was extremely low (<5%) in SOD1G93A mice. However, in the rare cases where the engrafted ChR2+ motor neurons did survive, we observed robust intramuscular axon growth and partial reinnervation of NMJs, even at very late-stage disease (Figure 1A-C and Video 1). Furthermore, acute optical stimulation of the engrafted ChR2+ motor neurons was able to induce tetanic contraction of the target muscle in these SOD1G93A mice (Figure 1D), although these were weak in magnitude. Interestingly, assessment of shorter timepoints revealed that engrafted ChR2+ motor neurons were capable of surviving for up to 14 days (data not shown). It is therefore likely that the poor long-term survival was due to host-vs-graft rejection of the engrafted cells, rather than disease-related toxicity. These findings suggest that if the engrafted motor neurons can evade the host immune response, they can functionally reinnervate target muscles for a therapeutically relevant timescale in this highly aggressive model of ALS.

Engrafted allogeneic ChR2+ motor neuron can survive and robustly reinnervate target muscles in SOD1G93A mice but occurs rarely.

(A) Confocal tile-scan of a longitudinal section of the triceps surae muscle from a 135d SOD1G93A mouse, 45d post-engraftment, showing a rare example of graft survival, in the absence of immunosuppression, with robust intramuscular axon projection. (B) Maximum intensity projection (MIP) images of a confocal z-stack through region of interest (dashed box in (A)) and (C), showing NMJs fully (asterisks) and partially (arrow) innervated by ChR2+ motor neuron axons, as well as innervated by endogenous (GFP/YFP-negative, ChAT-positive) motor axons (arrowhead) and fully denervated endplates (circle). (D) Representative recording (n = 1/3 positive responders) showing characteristically weak contractile responses to repetitive 20 Hz optical stimulation.

Tacrolimus (FK506) overcomes ESC-MN allograft rejection but inhibits muscle innervation

In an effort to enhance graft survival, we first tested the calcineurin inhibitor, tacrolimus (FK506), which is not only routinely used to facilitate solid organ allograft survival in humans but has also been reported to promote axon regeneration after nerve injury in rats Gold et al. (1995). Cohorts of wild-type and SOD1G93A mice (Supplementary Table 2) were treated with FK506 (5mg, kg-1, d-1; dose selected based on evidence of axon regeneration studies) from the time of intraneural ChR2+ motor neuron engraftment; graft survival and muscle reinnervation were then assessed 30-35d post-engraftment. Although FK506 did enable robust ChR2+ motor neuron allograft survival, at the graft site, in all animals examined (Supplementary Table 2), we identified three major problems with this immunosuppressant agent. Firstly, although the engrafted motor neurons were able to survive and project axons along peripheral nerve branches and within the targeted lower hindlimb muscles (Figure 2A), FK506 completely prevented muscle fibre reinnervation, evidenced by lack of response to acute ONS (data not shown) and complete absence of ChR2+/YFP+ NMJs (Figure 2B) in both SOD1G93A (n = 8) and nerve-ligated wild-type mice (n = 4). Secondly, exuberant growth of carry-over pluripotent stem cells led to intraneural tumour formation in most mice (Figure 2C and Supplementary Table 2), which resulted in overt hindlimb motor deficits (Figure 2D and Video 2). Thirdly, in partial agreement with previous reports in rats Aydin et al. (2004), FK506 prevented body mass increase and/or induced body mass decline in a subset (44.4%; n = 8/18) of SOD1G93A mice (Figure 2E and Figure 2-figure supplement 1A); this effect was less evident in wild-type mice (Figure 2-figure supplement 1B). Onset of body mass decline in B6.SOD1G93A mice is highly consistent and typically occurs at 115 days in Hayworth and Gonzalez-Lima (2009), indicating that, at this relatively high dose, FK506 may be preferentially toxic or may exacerbate disease phenotype in SOD1G93A mice.

FK506 enables long term survival of engrafted ChR2+ motor neurons but inhibits muscle reinnervation and exacerbates disease progression in SOD1G93A mice.

(A) Representative confocal image showing that GFP/YFP+ axons are able to project within intramuscular branches, following intraneural engraftment of ChR2+ motor neurons and daily immunosuppression with FK506; obtained from a 112d SOD1G93A mouse at 27d post-engraftment. (B) GFP/YFP+ axons fail to reinnervate NMJs despite the proximity of ChR2+ motor axon terminals to denervated endplates. (C) FK506-mediated immunosuppression permits exuberant growth of carry-over pluripotent stem cells that form intraneural tumours (**) within engrafted sciatic nerve (SN) branches; (D) these tumours cause severe movement impairment of the affected hindlimb. (E) FK506 caused body mass loss in a subset (8/18) of SOD1G93A mice, treated at different ages, that precedes onset of normal decline in this model (indicated by vertical dashed line at 115d).

Since peripheral neuropathy is a known adverse event associated with calcineurin inhibitors Arnold et al. (2013), it is possible that FK506 treatment alone may adversely affect endogenous or engrafted motor axons. Indeed, examination of the cross-sectional area distribution of total (ie. sensory and motor) and motor neuron axons in branches of the sciatic nerve in FK506-treated SOD1G93A and WT mice (Figure 3A) revealed a significant loss of axons that affected most axonal calibres in the tibial nerve in wild-type and SOD1G93A mice (Figure 3B and C). A more pronounced loss of total and motor axons, spanning medium to large sized axonal calibres, was observed in the common peroneal nerve branch (Figure 3D and E), which indicates that FK506 can not only exacerbate ongoing motor axon loss in SOD1G93A mice but can also induce motor axon loss even in wild-type mice. Importantly, these neuropathy-like effects appear to be specific to FK-506, since an alternative immunosuppressant, H57-597 mAb (discussed in detail below), did not significantly alter total or motor axon size distribution or total axon counts (Figure 3-figure supplement 2), compared to untreated SOD1G93A mice.

FK506 causes peripheral nerve axonopathy in SOD1G93A and wild-type mice.

(A) Representative examples of wild-type (top) and SOD1G93A (bottom) sciatic nerve transverse sections, showing common peroneal nerve (CPN) and tibial nerve (TN), labelled for total axons (βIII tubulin; green) and motor axons (choline acetyl transferase; ChAT; red); automated axon size distribution analysis of (B) total and (C) motor axon in the tibial nerve (TN); (D) total and (E) motor axons in the common peroneal nerve (CPN) in both wild-type and SOD1G93A mice. Data shown as mean; error bars = SEM; 2-way ANOVA analysis: *denotes p ≤0.05; ** denotes p ≤0.0002; *** denotes p ≤0.002; **** denotes p ≤0.00002; significance bars displayed in red highlight changes directly attributable to FK506, independent of genotype.

Since FK506 is known to suppress myoblast proliferation and differentiation Hong et al. (2002) and can cause rare cases of myopathy in humans Breil and Chariot (1999), it is also possible that the FK506-dependent prevention of muscle reinnervation by engrafted ChR2+ motor neurons is due to a muscle specific effect. In any case, these findings clearly show that FK506 is unsuitable as an immunosuppressant to support ChR2+ motor neuron allograft survival and, indeed, suggest that long-term administration at this relatively high dose should be avoided in ALS patients. Importantly, however, the complete protection of engrafted ChR2+ motor neurons by FK506 did confirm that host-vs-graft rejection was responsible for the poor long-term graft survival observed in the absence of immunosuppression, rather than disease related neurotoxicity.

T-cell modulatory immunosuppression confers graft survival and target muscle innervation

In light of the deleterious effects of FK506, and given our aim of conferring compatibility of allogeneic ChR2+ motor neurons as a universal cell therapy for ALS, we sought to identify a more specific form of immunosuppression that avoids the negative effects of FK506 yet supports long term graft survival. Therefore, we investigated the T-cell receptor-beta (TC-beta) targeting monoclonal antibody, mAb H57-597, which has previously been shown to effectively promote long-term heart allograft survival in mice Miyahara et al. (2012) In addition, since our findings indicated that immunosuppression results in a greater risk of tumour formation from carry-over pluripotent stem cells, differentiated ChR2+ motor neurons were also pre-treated with mitomycin-C (MMC; 2 μg/ml for 2 hrs) prior to engraftment, to eliminate tumorigenic cells Magown et al. (2016), to further enhance the translational potential. MMC-treated ChR2+ motor neurons were unilaterally engrafted into the tibial nerve of symptomatic SOD1G93A mice (aged 95.7 ±4.6 days) in conjunction with transient H57-597 mAb delivery (1mg, kg-1; i.p. injection at 0, 1-, 3-, 7- and 14-days post-engraftment). The extent of reinnervation and the ability to optically control the function of the triceps surae (TS) muscle group in the lower hindlimb of SOD1G93A mice was then assessed at late-stage disease (133 ±6.9 d; n = 12) by determining the physiological response of the reinnervated muscles to acute ONS of the engrafted motor neurons followed by histological analysis of muscle and nerve. Histological analysis confirmed that in mAb H57-597 treated animals, engrafted ChR2+ motor neurons were present in all recipient SOD1G93A mice (n > 84 to date). Importantly, there was significant axonal projection within intramuscular nerve branches and robust reinnervation of muscle fibres in the targeted TS muscle (Figure 4A and 4B; Video 3 and 4). As we previously reported in a wild-type mouse nerve ligation model Bryson et al. (2014), some de novo NMJs in SOD1G93A mice exhibited signs of immaturity, including poly-innervation (Figure 4C), as well as collateral and terminal sprouting of motor axons (Figure 4D). Since the peripherally-engrafted reinnervating motor neurons are inactive during the post-engraftment period and progressive muscle atrophy is ongoing, only 10.5% of endplates are innervated by engrafted ChR2+ motor neurons (Figure 4E). Importantly, these reinnervated end-plates are functional, since acute in vivo ONS of the engrafted ChR2+ motor neurons in the exposed sciatic nerve of late-stage disease SOD1G93A mice at 133 ±7.2 days of age (37.7 ±5.1 days post-engraftment; n = 12), induced positive contractile responses in all animals, although the amplitude of the maximal contractile force was still weak (0.8 ± 0.2 g; n = 11).

Transient H57-597 mAb treatment confers complete ChR2+ motor neuron allograft survival and, importantly, allows robust triceps surae muscle reinnervation up until late-stage disease in SOD1G93A mice.

(A) 3-D reconstruction of 67 individual tile-scans, acquired from serial sections from an entire triceps surae muscle, from a 135d SOD1G93A mouse following engraftment of ChR2+ motor neurons at 95d and H57-597 treatment, showing the full extent of axonal projection throughout the whole muscle; see also Video 1. (B) A high-resolution maximum intensity projection (MIP) image of a confocal z-stack, revealing multiple NMJs innervated to varying extents (α-bungarotoxin; α-BTx; red)) by YFP+ engrafted motor neuron axons (green) labelled for choline-acetyl transferase (ChAT; blue – note; overexposure of blue channel enables visualization of muscle fibres). (C) MIP image of a confocal z-stack showing an example of a partially innervated NMJ. (D) MIP image of a confocal z-stack showing an example of a fully innervated NMJ; note, poly-innervation, shown in (C), and terminal sprouting, shown in (D) which are signs of immaturity. (E) Automated quantification of total endplate number (count = 3,482; labelled with α-BTx) and manual counts of endplates (count = 364) that showed YFP colocalization, indicating innervation, from the same muscle.

Motor neuron subtype identity does not affect response to optical stimulation

In an effort to increase the amplitude of the contractile response of the target muscle to optical stimulation we next tested whether engraftment of motor neurons with a fast-firing subtype identity may be more suitable than engraftment of predominantly slow-firing motor neurons by using alternative differentiation protocols.32 The stronger, predominantly fast-twitch, gastrocnemius component of the TS muscle group is usually innervated by fast-firing/fast-fatigable (FF) motor neurons, which are known to have the capacity to innervate many more individual muscle fibres per motor unit than slow-firing motor neurons, which normally innervate a much smaller number of weaker, slow-twitch muscle fibres, predominantly within the soleus and plantaris regions of the triceps surae. MMC-treated ChR2+ motor neurons, differentiated to yield FF subtype identity motor neurons, were engrafted into the tibial nerve of (106 ±7.2 days) SOD1G93A mice, in combination with transient H57-597 mAb treatment. Maximum isometric muscle contraction of the TS muscle in response to acute optical stimuli was then determined at the same age (133.9 ±7.2 days, n = 13) as previous grafts of predominantly slow-firing motor neurons (133 ±6.9 days, n = 11). This physiological analysis revealed that the motor neuron subtype identity did not significantly affect amplitude of the muscle response to acute optical stimulation (Figure 4-figure supplement 1A) and that the maximum contractile force elicited by ONS remained extremely weak compared to supra-maximal ENS (Figure 4-figure supplement 1B). This result implies that, unlike during normal neuromuscular development, motor neuron subtype identity is not an important determinant of muscle fibre innervation in the mixed fibre type triceps surae muscle. This finding has significance for future clinical translation, since only a single subtype of motor neuron may be required to innervate a variety of different muscles. Only MMC-treated motor neurons with a slow-firing medial motor column identity were used in the subsequent experiments reported here. Since modification of motor neuron subtype identity did not enhance contractile response, our next challenge was to identify an effective method to enhance reinnervation and force generating capacity of the targeted muscle in response to optical stimulation.

Optical stimulation training significantly enhances target muscle force generation

Spinal motor circuit development Milner and Landmesser (1999); Li et al. (2005) and NMJ formation/maintenance Sanes and Lichtman (1999) are known to be activity-dependent processes, thus, without regular stimulation, although peripherally engrafted ChR2+ motor neurons may survive, they are unlikely to form mature NMJs and will therefore have little effect on declining muscle function and atrophy in SOD1G93A mice. Moreover, paralysis and atrophy of affected muscles proceed unchecked in the SOD1G93A mouse model of ALS. We therefore investigated whether regular activation of the engrafted ChR2+ motor neurons, in conjunction with H57-597 mAb treatment, could enhance NMJ maturation and maximize the ability of target muscles to generate contractile force in response to acute optical stimulation in late-stage SOD1G93A mice. To do this, we adapted a wireless, fully implantable optical stimulation system,33 in order to implement a daily optical stimulation training regimen for engrafted mice. First, we modified the PCB design to simplify assembly (Figure 5-figure supplement 2A, implemented a new encapsulation method to ensure long term survival of the devices after implantation (Figure 5-figure supplement 2B) and incorporated an RF signal switch and pulse controller to deliver precisely timed RF pulses to power a 470nm light emitting diode (LED) connected to the implantable device (Figure 5-figure supplement 2C).

The modified optical stimulation devices were then surgically implanted in SOD1G93A mice concomitantly with intraneural engraftment of ChR2+ motor neurons, with the trailing LED positioned in close apposition to the graft site (Figure 5A). Commencing at 14 d post-engraftment, to allow growing ChR2+ motor axons sufficient time to reach the target muscle, the mice were transferred to a custom built chamber located above a resonance-frequency induction cavity for 1 hr per day, in order to undertake optical stimulation training (OST; Figure 5B and Video 5), using a bespoke pulse pattern that was empirically determined to elicit maximum contraction (Figure 5-figure supplements 3, 4 and 5), followed by a 2 s rest interval. Following daily OST in engrafted SOD1G93A mice for 21 days, isometric muscle tension physiology was performed at late-stage disease to determine the maximal contractile force of the TS muscle elicited by acute ONS of the exposed sciatic nerve graft site. In confirmation of our hypothesis, there was a highly significant, 9.4-fold, increase in the maximal tetanic force (7.5 ±0.94 g versus 0.8 ±0.2 g; p = ≤0.000001) elicited by ONS in the engrafted OST group of SOD1G93A mice at late-stage disease, compared to age-matched untrained SOD1G93A mice (132.4 ±6.8 days versus 133 ±6.9 days; n = 7 and 11, respectively; Figure 5C-G), although OST did not alter contractile rate characteristics (Figure 5-figure supplement 6A-C). Moreover, quantitative comparison of the maximum force elicited by ONS compared to ENS of TS muscles in late-stage SOD1G93A mice (Figure 5H), showed that in mice that underwent OST, acute ONS elicits up to 22.7% (± 4.5) of total residual muscle force produced by supra-maximal ENS, which activates both endogenous and engrafted motor neurons (Figure 5I), in contrast to only 1.46% (± 0.18) in untrained SOD1G93A mice (Figure 5-figure supplement 7A and 7B). This represents a remarkable >13-fold improvement in force generation. In engrafted SOD1G93A mice that did not undergo OST, the extremely weak twitch contractions in response to ONS precluded interrogation of individual motor unit force values and determination of motor units number estimates (MUNE) in most mice. In contrast, in SOD1G93A mice that underwent OST, the significantly increased contractile response to ONS enabled clear separation of individual motor unit values (Figure 5J), enabling MUNE values to be determined (Figure 5-figure supplement 6D). Furthermore, as we previously reported in nerve-ligated WT mice,22 delivery of repetitive ONS pulses (250 ms bursts of 20 Hz illumination, every 1 s, for 180 s duration) to induce rhythmic, submaximal contraction of the TS muscle did not induce rapid muscle fatigue, whereas equivalent pulses of ENS stimulation of the contralateral TS muscle did result in rapid muscle fatigue (Figure 5K). This observation has significant implications for the ability to control repetitive motor functions in ALS patients.

Daily optical stimulation training (OST) of post-symptom onset SOD1G93A mice engrafted with ChR2+ motor neurons, significantly enhances contractile response to optical stimulation.

(A) Schematic indicating intraneural engraftment site in distal tibial nerve and reinnervated triceps surae (TS) muscle, along with stimulation device (top inset) implantation site and subcutaneous LED position; +4 div MMC-treated ChR2+ motor neurons express YFP (green; lower inset box); experimental timescale is shown below. (B) Still frame (taken from Video 5) showing daily OST. Representative isometric muscle tension physiology recordings from the TS muscle in response to specified pulses of ONS in untrained (C) and +OST (D) late-stage SOD1G93A mice, along with ENS response from the same muscle (E). (F) Delivery of an optimized pulse pattern elicits maximal response to ONS that can be used to finely control repetitive contractions; (G) Dashed box is shown at higher temporal resolution to indicate square-wave TTL pulse pattern (that drives LED stimulator) and an individual tetanic contraction. (H) Quantification of maximum contractile responses to indicated pulse patterns of acute ONS shows a highly significant improvement in force generation in late-stage SOD1G93A mice that underwent OST versus untrained controls (dashed horizontal line indicates maximum value from our previous study in nerve ligated WT mice). (I) The proportion of total muscle capacity (determined by supramaximal ENS minus ONS value) elicited by acute ONS is also significantly higher following OST; data represent mean ± SEM. (J) Motor unit number estimate (MUNE) traces obtained from a representative late-stage SOD1G93A mouse, following OST, in response to ipsilateral ONS and ENS, along with contralateral ENS (note, different scales). (K) Fatigue trace recordings comparing ipsilateral ONS (top) and contralateral ENS (bottom) in the same late-stage SOD1G93A mouse, in response to 250ms 20Hz pulses, repeated every 1s for 180s.

Optical stimulation training prevents atrophy of reinnervated muscle fibres

Finally, having established that optical stimulation training significantly enhances the maximal force elicited by optical stimulation of engrafted ChR2+ motor neurons in late-stage SOD1G93A mice, we examined whether long-term optical stimulation training could also prevent atrophy of reinnervated muscle fibres. Since NMJs comprise an extremely small volume of the entire muscle, high-resolution 3D imaging of the entire muscle to determine muscle fibre innervation status and fibre diameter information is not feasible. Therefore, we developed a novel technique, termed digital Cross-sectional area Analysis from Longitudinal Muscle Sections (dCALMS), in order to assess these properties. This involved 3D reconstruction and analysis of regions of interest, obtained from 30μm thick longitudinal TS muscle sections (Figure 6A) from ChR2+ motor neuron engrafted, late-stage SOD1G93A mice that had undergone OST. Each region contained at least one NMJ innervated by a ChR2+ motor neuron, along with randomly captured neighbouring fibres (Figure 6B, Figure 6-figure supplement 8 and Video 6). The 3D reconstructions were then digitally re-sliced in the transverse orientation (Figure 6C, top panel), in order to obtain data on muscle fibre cross-sectional area (CSA) using a semi-automated process (Figure 6C, lower panel). The dCALMS analysis revealed that the average CSA of muscle fibres innervated by engrafted ChR2+ motor neurons were similar in size to fibres still innervated by residual endogenous motor neurons (922.3 versus 1018.3 μm2; p = ≤0.85; Figure 6D). Importantly, the CSA of muscle fibres innervated by ChR2+ motor neurons was significantly greater than fibres with completely denervated endplates (average CSA = 525.4 μm2; p = ≤0.00001) or fibres whose innervation status could not be determined (average CSA = 668.8 μm2; p ≤0.0003), since the endplate was outside the scanned region of interest.

Optical stimulation training prevents atrophy of muscle fibres that have been reinnervated by ChR2+ motor neurons in late-stage SOD1G93A mice.

(A) Confocal tile-scan showing a single longitudinal section through the triceps surae muscle of 135d SOD1G93A mouse (35d post-engraftment), following daily OST; endplates (labelled with αBTx) innervated by GFP/YFP+ engrafted motor neurons are evident throughout the whole muscle group, including fast-twitch medial gastrocnemius (MG) and lateral gastrocnemius (LG) muscles and the slow-twitch soleus (Sol) muscle. (B) Representative top-down maximum intensity projection (MIP) view of a confocal z-stack through a 30μm longitudinal section obtained from the same mouse; including side-on (B’) and end-to-end (B”) MIP views of the same z-stack; a de novo NMJ, innervated by a ChR2+ motor neuron (asterisk), is indicated on a muscle fibre that also has a denervated endplate (arrow), along with another muscle fibre still innervated by an endogenous choline-acetyltransferase (ChAT+) positive, GFP-negative motor neuron (arrowhead). (C) A digital slice (top) through the 3-D z-stack obtained at the y-axis plane indicated by dashed line in (B) and colorized masks delineating individual muscle fibres (bottom); see Video 6. (D) Average cross-sectional area of individual muscle fibres with an innervation status that was unconfirmed (117 fibres), innervated by GFP+ motor neurons (62 fibres), denervated (28 fibres), or innervated by endogenous motor axons (13 fibres); n = 3 late-stage engrafted SOD1G93A mice that had undergone OST; Data shown as mean; error bars = SEM; one-way ANOVA with Tukey’s post hoc correction: *denotes p ≤0.05; ** denotes p ≤0.0002; *** denotes p ≤0.002; **** denotes p ≤0.00002.

Discussion

This study shows for the first time that replacement stem cell-derived motor neurons can robustly and reliably reinnervate target muscles in the highly aggressive SOD1G93A mouse model of ALS, even when engrafted after onset of overt symptoms; moreover, the restored innervation can be maintained even until extremely late-stage disease. Furthermore, our findings suggest that engrafted ChR2+ motor neurons can not only provide an interface to safely and selectively control the function of targeted muscle but, also, that regular optical stimulation training (OST) can be used to: i) reinforce connectivity between engrafted motor neurons and muscle fibres; ii) significantly enhance the maximal force elicited by optical stimulation of the targeted muscle; and iii) prevent atrophy of muscle fibres that are reinnervated by engrafted motor neurons. The highly significant improvements in muscle innervation, atrophy prevention and maximum contractile force, as a result of the daily OST regimen, confirms that stimulation-induced activity is necessary to maximize connectivity between engrafted motor neurons and their target muscles.

The prevailing view in the ALS field is that spinal motor neuron pathology first manifests at the nerve terminal and that muscle fibres are likely to actively contribute to degeneration of endogenous motor neurons in ALS Scaricamazza et al. (2021). Our findings clearly demonstrate that affected muscles, even in the highly aggressive SOD1G93A ALS model, remain receptive to reinnervation by healthy engrafted motor neurons, even until late-stage disease. Moreover, once target muscles have been reinnervated, this approach enables implementation of a safe muscle training/exercise regimen that can be used to preserve muscle integrity and prevent irreversible muscle wasting that otherwise occurs as a result of progressive neurodegeneration in ALS Mora (1989). Since skeletal muscles are not simply biomechanical actuators, but have complex functions in overall metabolic homeostasis, thermoregulation, venous return and maintenance of blood volume, the ability to prevent muscle atrophy using OST, is likely, by itself, to have major health benefits for ALS patients. In contrast, the use of electrical nerve stimulation (ENS) to control muscle function or, indeed, exercise programs is likely to accelerate degeneration of remaining motor axon terminals Guimarães-Costa et al. (2019), and is therefore unlikely to be safe. Unlike ENS, the highly selective nature of ONS does not activate or interfere with endogenous motor neuron function and due to its ability to recruit motor units in physiological order, ONS has the added significant benefit of avoiding rapid muscle fatigue Llewellyn et al. (2010). The ability to combat muscle atrophy, using OST, could extend the ability of targeted to execute functionally useful movements, potentially indefinitely in ALS patients.

Since differentiation methods that yield either fast-firing or slow-firing motor neurons did not appear to affect the ability of engrafted motor neurons to innervate the mixed fibre type TS muscle group in SOD1G93A mice, motor neuron subtype identity appears to be redundant in this case. Therefore, a single type of motor neuron, produced at scale, could potentially be used to target a large number of different muscles in each patient. This has advantages in terms of simplifying the regulatory approval process, since the donor motor neurons would be produced in exactly the same way, irrespective of the graft site or recipient.

Indeed, another key finding of this study that could streamline future translation of this therapeutic strategy, is the identification of a TCR-β targeting antibody, H57-597 mAb, as an effective mediator of allograft survival. Existing T-cell targeting monoclonal antibodies, such as OKT-3, have been clinically approved Page et al. (2013) and, importantly, this form of immunosuppression overcomes severe adverse effects that we observed with the commonly used CNI-based immunosuppressant, tacrolimus (FK506). Our data shows that H57-597 mAb treatment was well tolerated during transient administration from symptom-onset up until late-stage disease in SOD1G93A mice, however, the aggressive disease progression in this model precludes investigation of longer-term effects and it remains to be seen how long-term TCR-β-based immunosuppression may be tolerated in ALS patients. A similar immunosuppression regimen involving a CD25 (IL2) targeting monoclonal antibody, Basiliximab, along with glucocorticoid and tacrolimus maintenance therapy, has been shown to be safe but poorly tolerated in ALS patients, as an independent investigative approach Fournier et al. (2018), as well as part of a separate clinical trial assessing intraspinal grafts of neural precursor cells in ALS patients Mazzini et al. (2019). Therefore, immunosuppression per se, is not an impediment to this strategy.

The use of allogeneic donor cells, with a safe and effective immunosuppression regimen, means that a future cell therapy could potentially be universally compatible with all ALS patients, which would significantly reduce costs and simplify the regulatory approval process, compared to individually tailored autologous cell grafts. Of course, further studies will be required to ensure that human-compatible, induced pluripotent stem cell (iPSC)-derived donor motor neurons are able to function in the same manner as allogeneic mouse ESC-derived motor neurons. The generation of HLA-matched super donor hiPSC lines may further mitigate the need for immunosuppression Turner et al. (2013), however, immunogenicity of the ChR2 protein could mean that some form of immunosuppression may be necessary for any optogenetic therapy in the peripheral nervous system, including viral delivery Maimon et al. (2018).

The anatomical separation of specific nerve branches in humans means that this approach could be used to target and independently control large numbers of muscles in each individual ALS patient. However, it will be first necessary to demonstrate safety and efficacy in a relatively low risk muscle to restore a simple motor function. For example, the common peroneal nerve is highly accessible from a neurosurgical perspective and existing ENS devices, developed to correct foot drop for other neurological disorders Hausmann et al. (2015), could be readily adapted to assist ambulatory function in early-stage ALS patients. In the longer term, existing multichannel ENS devices, which have been developed to control more complex ADLs in high-level spinal cord injury (SCI) patients Memberg et al. (2014), could also be adapted into a minimally invasive, transcutaneous optical stimulation device Maimon et al. (2017). This combinatorial therapeutic strategy, comprising allogeneic donor cells, an effective immunosuppression regimen and optical stimulation device, is highly compatible with the rapidly evolving field of brain computer interface (BCI) technology. BCI could be used to decode a paralyzed patient’s intention to perform a given movement in order to control the activity of the engrafted motor neurons, which provide the necessary interface to execute the intended movement via a wearable optical stimulation device. This novel approach would entirely bypass the severe damage that occurs throughout the entire CNS in ALS patients and enable autologous control of movement. Moreover, this strategy also has broad utility for a wide range of other neurogenic causes of paralysis, such as spinal cord injury and stroke.

Although the findings of this study clearly demonstrate that our combinatorial cell therapy approach is effective in a highly aggressive mouse model of ALS, further investigation is required in order to confirm that the strategy can be applied to alternate model of ALS, with a longer lifespan, in order to fully explore the long-term efficacy of the approach, particularly in terms of chronic allograft survival using a transient immunosuppression approach. It is possible that some form of maintenance therapy may also be required to confer long term graft survival. Of course, the biggest challenge will be to demonstrate that human optogenetically-modified motor neurons, derived from either induced pluripotent stem cells (iPSCs) or human ESCs, are capable of reinnervating target muscles in the same manner as we have demonstrated for mESC-derived motor neurons. It will also be necessary to scale this cell therapy strategy up, using larger animal models that more accurately recapitulate human-scale anatomy.

Despite these remaining challenges, the findings of this study provide strong support for this novel cell therapy, which, if successful, could finally begin to deliver major health benefits for ALS patients.

Methods and Materials

Detailed methods are provided in the Supplementary material.

mESC motor neuron differentiation

The Channelrhodopsin2-YFP expressing mESCs (Clone C9G) used in this study were generated as previously describedBryson et al. (2014). Motor neuron differentiation was performed according to a standard protocol developed by Wichterle et al Wichterle et al. (2002) for production of predominantly slow-firing medial motor column (MMC) identity motor neurons and an updated “caudalized-ventralized” (CV) protocol, also developed by the Wichterle group Peljto et al. (2010), that produces a higher proportion of motor neurons with fast-firing properties. Following differentiation on Day 5 (or Day 7 for the C-V protocol), embryoid bodies (EBs) containing differentiated motor neurons were dissociated and total cells were resuspended in PBS at a concentration of 50,000 cell/μl. Where indicated, mitomycin-C (1μg/μl) was added to the EBs for 2 hrs prior to dissociation. Nile Blue A (0.0002% final concentration) was added to the cell suspension and the cells were kept on ice until engraftment. Single nucleotide polymorphism (SNP) analysis of mESC lines (Clone C9G, HBG3 and a C57BL/6J mESC line for reference) was carried out by Charles River Laboratories. See the Supplementary materials for full details 1.

Mice

All procedures and experiments involving animals were carried out under License from the UK Home Office in accordance with the Animals (Scientific Procedures) Act 1986 (Amended Regulations 2012), following ethical approval from the UCL Queen Square Institute of Neurology Animal Welfare Ethical Review Body (AWERB), and in accordance with the ARRIVE guidelines. B6.Cg-Tg(SOD1*G93A)1Gur/J mice (The Jackson Laboratory, stock number 004435) were bred specifically for this study by mating presymptomatic male transgene carriers with congenic C57BL/6J females (Charles River Laboratories)1.

Intraneural engraftment of mESC-derived ChR2+ motor neurons

Surgical engraftment of ChR2+ motor neurons was performed under aseptic conditions, as previously described Bryson et al. (2014). Briefly, 1μl of dissociated EB cell suspension (50,000 cells) was injected into the tibial nerve close to the trifurcation point of the sciatic nerve, using a 5μl Hamilton syringe equipped with a customized 33G needle. Where indicated, an implantable optical stimulation device was inserted through the same surgical incision and positioned subcutaneously under the skin on the back; the trailing LED was fixed with sutures to the overlying muscles at the graft site during wound closure. Immunosuppression, as indicated, was initiated at the time of surgical engraftment. See supplementary materials for full details1.

Implantable optical stimulation devices and power transmission system

Optical stimulation devices were largely produced as described by Montgomery et al. Montgomery et al. (2015), with minor, but essential, modifications to the PCB design and encapsulation method. Similarly, the power transmission system used to activate the implanted LED devices was also largely as described by Montgomery et al. Montgomery et al. (2015), however, a Solid State Switch (MiniCircuits; ZX80-DR230-S+), controlled by a USB-TTL Interface (Prizmatix), was used to generate specific optical stimulation patterns. Engrafted SOD1G93A mice that underwent daily optical stimulation training were placed in the stimulation chamber for 1 hr/day from post-engraftment day 14 until termination of the experiment at late-stage disease (132.4 ±6.8 days). See the Supplementary materials for full details1.

Isometric muscle tension physiological analysis

At the experimental end-point, engrafted SOD1G93A mice underwent isometric muscle tension physiology, in order to accurately determine the contractile properties of the triceps surae muscle in response to acute optical stimulation, as previously described Bryson et al. (2014), with the following modifications: a PowerLab 4/30 stimulation and recording system (AD Instruments) was used to deliver bespoke electrical stimulation signals, either as direct constant voltage pulses applied to the nerve via bipolar electrodes for electrical nerve stimulation (ENS), or used as a 5V TTL signal to activate a 470nm LED light-source (CoolLED; pe100), delivered to the exposed sciatic nerve via a liquid lightguide for optical nerve stimulation (ONS); stimulation program available on request. LabChart software (AD Instruments) was used for automated data analysis of contractile parameters. See the Supplementary materials for full details1.

Nerve and muscle histology and image analysis

See Supplementary materials for full details, including automated motor/sensory axon CSA analysis1, axon counts and innervation analysis method1. The digital CSA analysis of longitudinal muscle sections (dCALMS) method is also described in full in the Supplementary materials1.

Quantification and statistical analysis

Sample sizes

The number of mice (n) is provided in the figures and/or figure legends; also see Supplementary Tables 2 and 3.

Statistical analysis

All data are presented as mean ± SEM unless otherwise indicated. GraphPad Prism 9 (Prism) was used for statistical analyses. No out-liners or data points were eliminated. Differences between two groups were assessed using multiple two-tailed unpaired t tests. Differences between more than two groups were assessed by using one-way or two-way analysis of variance (ANOVA) with multiple comparison correction, or mixed model effects analysis, as stated in the figure legends. Significance was defined as *P ≤0.05, **P ≤0.01, ***P ≤0.001, or ****P ≤0.0001. See supplementary Methods for further details 1.

Data availability

The data that support the findings of this study are available from the corresponding authors, upon reasonable request.

eLife Assessment

This fundamental study presents a method to restore muscle innervations in ALS mouse models using optogenetics. It is convincing that embryonic stem cell derived motor neurons can be transplanted into and applied to reinnervate the muscles in an ALS mouse model. The work will be of broad interest to researchers and medical biologists to develop new strategies for the treatment of neurodegenerative disorders resulting from denervated skeletal muscles.

Reviewer #1 (Public Review):

Amyotrophic lateral sclerosis (ALS) is a neurodegenerative disorder leading to the loss of innervation of skeletal muscles, caused by the dysfunction and eventual death of lower motor neurons. A variety of approaches have been taken to treat this disease. With the exception of three drugs that modestly slow progression, most therapeutics have failed to provide benefit. Replacing lost motor neurons in the spinal cord with healthy cells is plagued by a number of challenges, including the toxic environment, inhibitory cues that prevent axon outgrowth to the periphery, and proper targeting of the axons to the correct muscle groups. These challenges seem to be well beyond our current technological approaches. Avoiding these challenges altogether, Bryson et al. seek to transplant the replacement motor neurons into the peripheral nerves, closer to their targets. The current manuscript addresses some of the challenges that will need to be overcome, such as immune rejection of the allograft and optimizing maturation of the neuromuscular junction.

Bryson et al. begin by examining the survival of mESC-derived motor neurons allografted into SOD1 mice. The motor neurons, made on a 129S1/SvImJ, were transplanted into the tibial nerve of SOD1 mice on a C57BL/6J background. Without immunosuppression, most cells were lost between 14 and 35 days, suggesting an immune response had eliminated them. Tacrolimus prevented cell loss, but it also inhibited innervation of the muscle. It also uncovered the tumorigenic potential of contaminating pluripotent cells. In contrast, immunosuppression using H57-597, an antibody targeting T-cell receptor beta, prevented graft rejection while permitting some innervation of muscle. Pretreatment of the cells with mitomycin-C eliminated pluripotent cells, preventing tumor formation. The authors noted that this combination only innervated ~10% of endplates, likely due to the fact that the implanted motor neurons are not active.

The authors then began the process of optimizing the cells themselves, using measurements taken in late-stage SOD1 mice. Fast-firing and slow-firing populations of neurons were first compared. Using optical stimulation, these two cell types appeared to be similar. The authors opted to use slow-firing neurons in the subsequent experiments. Recognizing that neuromuscular junction (NMJ) innervation and maintenance are dependent on motor neuron activity, implantable optical stimulators were also evaluated. 14 days after transplanting the cells, optical stimulation training was initiated for one hour each day. This training led to a nearly 13-fold increase in force generation, although this still remained well below the force generated by electrical stimulation. The enhanced innervation also prevented the atrophy of muscle fibers caused by denervation.

Overall, the data for the function of the implanted cells are convincing. The dCALMS technique that the authors have developed is quite interesting and will likely be applicable to analyze muscles for other therapeutics. The identification of calcineurin inhibitors as inhibitors of reinnervation will also be important for the development of other cell-based therapeutics for ALS.

However, there are some issues that should be addressed. These include some common misconceptions about ALS. While ALS is split into familial and sporadic forms based on the presence or absence of a family history of the disease, mutations in the known ALS-associated genes are found in both forms. The authors also state that exercise programs are likely to accelerate degeneration in ALS. This is incorrect. Moderate exercise is part of the current guidelines for treating ALS, and mouse studies have demonstrated a therapeutic effect of moderate exercise. Regarding the experimental design, there are some important details missing. The animals do not appear to have been operated on at the same age, and the criteria for when to perform the operation were not described. A similar problem exists for when the animals were determined to reach endpoint. The authors also do not seem to address a potential pitfall of this approach: acceleration of the disease process. Indeed, some of the data comparing the ipsilateral side to the contralateral side suggest that the implantation of the cells and/or the light source increase the denervation of the muscle. Finally, there is a fairly large difference between the motor output provided by optical stimulation relative to electrical stimulation. It is currently unclear what level needs to be reached to provide an effective response in the intact animal. Thus, it is difficult to determine if the level of reinnervation that this study has achieved will be sufficient to improve a patient's quality of life.

Reviewer #2 (Public Review):

The authors provide convincing evidence that optogenetic stimulation of ChR2-expressing motor neurons implanted in muscles effectively restores innervation of severely affected skeletal muscles in the aggressive SOD1 mouse model of ALS, and conclude that this method can be applied to selectively control the function of implicated muscles. This was supported by convincing data presented in the paper.

This is an interesting paper providing new/improved optogenetic methods to restore or improve muscle strength in ALS. In general, it is of high significance in both the techniques and concept, and the paper was well written. The evidence supporting the conclusions is convincing, with rigorous muscle tension physiological analysis, and nerve and muscle histology and image analysis. The work will be of broad interest to medical biologists on muscle disorders.

One weak point is that proper control experiments were not clearly presented - these could be shown in the paper. For example, one control experiment with only YFP but no ChR2 expression with optogenetic stimulation should be performed, following similar procedures and analysis applied to the ChR2-transduced animals.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Recommendations For The Authors):

Major concerns:

1. In lines 41-43, there seems to be some confusion about the terminology regarding "sporadic ALS". ALS is subdivided into familial and sporadic forms. Familial ALS simply indicates that the patient has a family history of ALS and presumably has a genetic predisposition for developing this disease. In many families, the identity of the mutation remains unknown. Sporadic ALS patients do not have a family history of this disease. However, this does not imply that they lack mutations that caused disease. In fact, 5-10% of these patients have the hexanucleotide repeat expansion in C9orf72. This mutation is also found in about 40% of familial ALS cases.

We have now amended the manuscript to be more accurate in our description of underlying genetics of ALS. This changes to this section are as follows:

Lines 39-47:

"...The median survival time in ALS, from initial onset of symptoms to death, typically as a result of respiratory complications, is only 20-48 months Chiò et al. (2009) and ALS has an estimated global mortality of 30,000 patients per year Mathis et al. (2019).

ALS is typically classified into either familial (fALS) or sporadic (sALS) forms of the disease, based on whether or not patients have an identified family history of the disease; between 5-10% of total ALS cases fall into the former category, fALS, with the remaining 90-95% consisting of sALS cases Mathis et al. (2019). To date, over 20 monogenic mutations that cause ALS have been identified, however these still only account for 45% of fALS cases and only 7% of sALS cases Mejzini et al. (2019)..."

1. In Fig. 4-supplement 1, 7DD and 5DD are not defined. I assume one is the fast-firing and one is the slow-firing motor neurons. I am also a bit confused as to why the 5DD neurons produce greater muscle force than the 7DD neurons when electrically stimulated. It seems to suggest that there is some difference between the two types of neurons or the groups of mice used to test them.

We have now defined these terms and the amended figure legend now reads as follows:

"(A) Fast-firing motor neurons (produced using a 7-day differentiation protocol thus labelled as “7DD”) or slow-firing ChR2+ motor neurons (produced using a 5-day differentiation protocol thus labelled as “5DD”) were engrafted in age matched SOD1G93A mice… Our expectation was that fast-firing motor neurons, which normally innervate larger numbers (>100) of stronger fast-twitch muscle fibres per motor unit would elicit significantly greater contractile force when optically stimulated, compared to slow-firing motor neurons that innervate small numbers (<10) of weaker, slow-twitch muscle fibres per motor unit. Surprisingly, our data did not show any difference when using grafts consisting of fast-firing motor neurons, versus slow-firing motor neurons, at least in response to optical stimulation. The factors underlying this surprising result, and the apparent discrepancy between electrically-evoked muscle contractions in nerves that had bene engrafted with either fast or slow firing motor neurons, are likely to be highly complex; we hope to further explore this as part of a separate follow up study."

1. Along those lines, do these two subpopulations of motor neurons innervate the same set of muscle fibers? More generally, are certain types of muscle fibers preferentially innervated by this approach? Answering these questions could point to additional ways to enhance the effectiveness of this treatment approach. This should be discussed.

This point is partially addressed in our response to Point 2 above, but to further extrapolate: certainly, the phenotype of individual muscle fibres is largely dictated by the firing properties of the motor neuron that innervates it. Slow-twitch muscle fibres tend to produce less contractile force but are more fatigue resistant, whereas fast-twitch muscle fibres produce more force but fatigue rapidly. There is evidence that expression of the chemorepellent molecule ephrin-A3 prevents the inappropriate innervation of slow-twitch muscle fibres by fast-firing motor neurons, which express the cognate receptor EphA8 [PMID: 26644518]. Importantly, fast-firing motor neurons are preferentially susceptible to disease mechanisms in ALS and the fast-twitch muscle fibres that they innervate are therefore more likely to undergo denervation and atrophy. Surprisingly, in this study we clearly show that grafts consisting of slow-firing motor neurons are able to innervate all regions of the triceps surae muscle group, including the normally exclusively fast-twitch superficial regions of the gastrocnemius and the exclusively slow-twitch soleus muscle. This finding strongly suggests that the normal developmental pairing of motor neuron and muscle fibre properties is not essential in this therapeutic context. Indeed, the use of more disease-resistant slow-firing motor neurons may provide some advantages. Again, we hope to be able to further explore this relationship in forthcoming follow-up studies.

1. The authors state that exercise programs are likely to accelerate disease progression. This is not supported by the current body of clinical data. In fact, current guidelines are for moderate (not strenuous) exercise, and mouse studies have demonstrated a protective effect of moderate exercise on disease progression.

We apologise for the lack of clarity on this point, as it was not our intention to imply that voluntary exercise accelerates disease progression. We have now amended the manuscript to specify “ENS-based exercise programs” to avoid any confusion.

1. It is unclear what the experimental endpoint is. Page 25 defines it as 135 days of age, but ranges are given the figure legends, suggesting that some other criteria were used. It also seems unclear at what determined the age at which each animal was treated since they were also not treated at the same age.

We hope that our response in the Public Reviews section above has fully addressed this point.

1. I am a little confused by Figure 5 - figure supplement 5, panel D. Why do the authors give specific p-values here but not in the other panels? The sample sizes in D are very low, in some cases with only 1 animal in a group, and performing statistical tests under these conditions seems futile. The statistical power is nearly zero.

For the purposes of consistency, we have now replaced the specific p-values in panel D with “ns”. The low n-values for the MUNE analysis data is due to the extremely difficult nature of identifying the contribution of individual motor units to the total muscle contractile response, when the maximal muscle force is extremely weak. In the absence of optical stimulation training, the extremely weak force elicited by acute optical stimulation precluded our ability to separate out the contribution of individual motor units and, often, in animals where this was not possible, we did not always perform electrically-evoked MUNE analysis. Unfortunately, we are not currently in a position to increase the n-values for this component of the study. Our ongoing research to enhance the amplitude of the muscle response to optical stimulation will hopefully help to more clearly address this in the future.

1. One concern about this approach is that the procedure could accelerate the denervation of the target muscle. Figure 5 - figure supplement 6, panel B, indicates a significant reduction in force on the ipsilateral side relative to the contralateral side, at least under electrical stimulation of the nerve. This would be consistent with the hypothesis that the procedure does enhance disease progression in the treated limb. Is there a reduction in voluntary motor activity in these animals, such as in grip strength or the position of the foot while walking?

We hope that this important point has been satisfactorily addressed in the Public Reviews section. Unfortunately, we did not undertake any behavioural analysis relating to voluntary motor function of the engrafted (or contralateral) hindlimbs, which may have provided useful data to address this point. As described above, the most likely explanation for this finding is due to physical nerve damage caused by the intraneural injection procedure; in our efforts to refine our strategy and move it towards clinical translation, we will take this into consideration in our future research.

1. Based on Fig. 6D, it seems that the vast majority of innervated NMJs at endpoint are innervated by cells from the graft. And yet, electrical stimulation evokes substantially greater muscle force. This may suggest that optical control of engrafted motor neurons will not yield enough force for routine tasks or that the few remaining endogenous motor neurons are much more effective at generating force. These potential limitations and ways to overcome them should be discussed.

There appears to be a slight misunderstanding, since our aim here was to sample a sufficiently powered number of motor end-plates innervated by YFP+ for statistical analysis. To do this we specifically chose regions of interest containing at least 1 YFP+ NMJ and the adjacent muscle fibres were included at random, whatever their innervation status. Had we sampled regions of interest at random, we would have been likely to capture only a very few YFP+ terminal as they occupy a very small volume of the total muscle section and the maximum scanning area for each high-resolution z-confocal stack is relatively small, so we feel that this selection was warranted.

Minor comments:

1. The donor mouse strain should be described as 129S1/SvImJ.

We have now corrected this.

1. The first time the supplementary figures show up in the manuscript, they seem to have two titles each, such as "Figure 1-figure supplement 1. (Figure 4 - figure supplement 1)". The second seems to be the correct one.

This was caused by an issue with the Latex template, which has now been resolved.

1. PCB is not defined the first time it is used (page 8, line 332).

We have now defined this term on first use: printed circuit board (PCB)

1. CNI is not defined in the text (page 12, line 432).

We have now defined this abbreviation at the first usage on Page 4, Line 158

1. Some of the fonts on the graphs are very small, such as Fig. 5J.

We have increased the font size as much as possible for Fig. 5.

1. Figure 6 - figure supplement 1 does not include a key to indicate which antigens are stains and which color refers to which antigen. This is also needed for the videos.

We have now included a key on this figure supplement to indicate the relevant antigens and stain and we have also done the same for the videos.

1. Video 5 seems to indicate that there is a dead zone in the back of the chamber. Does this raise any concerns about the consistency of training from animal to animal?

This is an extremely astute observation. However, the intermittent activation of the implantable LED devices is not due to a dead zone; rather, it is due to the orientation of the power receiving coil within the device and it’s alignment with the resonance frequency chamber that transmits the power to the device. As the animals move around, and particularly when they rear up, the power receiving coil occasionally becomes misaligned and fails to receive sufficient power to activate the LED. Since the pulses are delivered every 2 seconds, for 1 hour per day, we feel that the animals, on average, receive sufficient numbers of pulses to implement the training regimen. Indeed, we feel that the results speak for themselves.

Version 2

Abstract

Abstract

Breakdown of neuromuscular junctions (NMJs) is an early pathological hallmark of amyotrophic lateral sclerosis (ALS) that blocks neuromuscular transmission, leading to muscle weakness, paralysis and, ultimately, premature death. Currently, no therapies exist that can prevent progressive motor neuron degeneration, muscle denervation or paralysis in ALS. Here, we report important advances in the development of an optogenetic, neural replacement strategy that can effectively restore innervation of severely affected skeletal muscles in the aggressive SOD1G93A mouse model of ALS, thus providing an interface to selectively control the function of targeted muscles using optical stimulation. We also identify a specific approach to confer complete survival of allogeneic replacement motor neurons. Furthermore, we demonstrate that an optical stimulation training paradigm can prevent atrophy of reinnervated muscle fibers and results in a 10-fold increase in optically-evoked contractile force. Together, these advances pave the way for an assistive therapy that could benefit all ALS patients.

Introduction

The progressive degeneration of motor neurons that occurs in amyotrophic lateral sclerosis (ALS), the most common adult motor neuron disease (MND), affects almost all cellular components of the neuromuscular system, including cortical and spinal motor neurons, interneurons Crabé et al. (2020), glial cells Van Harten et al. (2021); Vahsen et al. (2021), as well as muscle Loeffer et al. (2016). Motor axons, contained within peripheral nerves, serve as the final common relay for transmission of motor signals that control all voluntary muscle contraction and movement. However, one of the earliest characteristic pathological features of ALS involves “die-back” of motor axon terminals Fischer et al. (2004) and breakdown of neuromuscular junctions (NMJs), the specialized synapses between motor axon terminals and muscle fibres. Effectively, this represents a single point of failure that permanently blocks motor signal transmission, irrespective of the condition of central motor circuits involved in coordination of movement signals. This results in an inexorable progression of muscle weakness, atrophy and, ultimately, complete paralysis, eventually leading to premature death. The median survival time in ALS, from initial onset of symptoms to death, typically as a result of respiratory complications, is only 20-48 months Chiò et al. (2009) and ALS has an estimated global mortality of 30,000 patients per year Mathis et al. (2019).

ALS is typically classified into either familial (fALS) or sporadic (sALS) forms of the disease, based on whether or not patients have an identified family history of the disease; between 5-10% of total ALS cases fall into the former category, fALS, with the remaining 90-95% consisting of sALS cases Mathis et al. (2019). To date, over 20 monogenic mutations that cause ALS have been identified, however these still only account for 45% of fALS cases and only 7% of sALS cases Mejzini et al. (2019). The downstream cellular and molecular pathomechanisms underlying neurodegeneration in ALS are extremely complex and include dysregulation of proteostasis, autophagy, RNA metabolism and axon transport, as well as excitotoxicity, oxidative stress and neuroinflammation Mejzini et al. (2019). Given the disparate causes and complex disease mechanisms, development of an effective therapy has proven extremely challenging and there are currently no effective treatments capable of arresting the progressive paralysis that occurs in ALS Brown and Al-Chalabi (2017). Even emergent gene therapy approaches, such as antisense oligonucleotides (ASOs) that have shown early promise in clinical trials Miller et al. (2013), are unlikely to benefit the majority of patients with sporadic ALS and will not replace degenerated motor neurons or restore motor function once it has been lost.

Similarly, efforts to develop cell therapy approaches for ALS have, so far, primarily been aimed at slowing the degeneration of motor neurons in the spinal cord by providing neurotrophic factor (NTF) support through intraspinal engraftment of foetally-derived neural stem cells (NSCs) Feldman et al. (2014); Goutman et al. (2018) or intrathecal delivery of autologous mesenchymal stem cells (MSCs) that are modified to overexpress NTFs Cudkowicz et al. (2022); Berry et al. (2019). Whilst these approaches were shown to significantly slow disease progression in transgenic animal models of ALS Xu et al. (2006); Yan et al. (2006), clinical trials in ALS patients have shown only a modest therapeutic effect in the case of intraspinal NSC grafts Glass et al. (2012) and no overall benefit of intrathecal MSC delivery Cudkowicz et al. (2022) Given the highly invasive surgical laminectomy required to engraft NSCs into the ventral horn of the spinal cord, only localized populations of motor neurons in the lumbar and/or cervical region were targeted, however, the surgical procedure was generally well tolerated and the approach was proven to be safe in clinical trials Feldman et al. (2014); Glass et al. (2016). Moreover, these clinical trials have provided evidence that specific muscle functions can be preserved for longer in some ALS patients Mazzini et al. (2019). Importantly, this pioneering approach provides a precedent for implementation of an allogeneic stem cell-based therapy and also shows that ALS patients can tolerate a 6 month period of immunosuppression Mazzini et al. (2015), which appears to be sufficient to confer long term survival of engrafted cells Tadesse et al. (2014). Nonetheless, it appears unlikely that this approach will be able to restore motor function once it has been lost, since the NSCs do not replace lost motor neurons, and any therapeutic effect has so far been shown to be transient Goutman et al. (2018); Mazzini et al. (2019). Therefore, there remains an urgent unmet need to develop novel therapies that can rescue muscle innervation and maintain muscle function in ALS patients.

We have previously demonstrated a novel proof-of-concept strategy to overcome muscle denervation and restore control of muscle contraction in a nerve injury model of muscle paralysis that could have major therapeutic value for restoring function of any targeted muscle or group of muscles in ALS patients Bryson et al. (2014). Briefly, optogenetically-modified replacement motor neurons, derived from murine embryonic stem cells (mESCs), were engrafted into distal branches of peripheral nerves supplying specific lower hindlimb flexor and extensor target muscles in wild-type mice that had undergone a nerve ligation injury. Our results showed that the engrafted motor neurons were able to project axons from the graft site to the target muscles where they formed de novo NMJs. Due to the ectopic location of the engrafted motor neurons, outside the central nervous system (CNS), they do not receive endogenous motor signals, and must therefore be exogenously activated. Expression of the blue-light sensitive channelrhodopsin-2 (ChR2) protein Nagel et al. (2003) in the engrafted motor neurons conferred the ability to selectively activate these engrafted neurons and thereby control the contractile function of the target muscle using acute optical stimulation Bryson et al. (2014). The aim of this neural replacement strategy is therefore to provide a biological interface capable of rendering any target muscle receptive to control signals transmitted by optical stimulation to engrafted motor neurons Bryson et al. (2016). Importantly, we have recently developed a prototype 64-channel stimulation and recording device capable of controlling multiple independent intraneural graft sites that could be used to elicit coordinated function of large numbers of muscles, in order to restore useful motor functions Liu et al. (2022).

This novel approach to restore control of paralyzed muscles in ALS patients, using a combination of cell replacement and optical stimulation, has several key advantages over existing cell replacement and electrical stimulation strategies, including: i) the ability to engraft motor neurons peripherally, in close proximity to targeted muscles, which greatly accelerates the rate of reinnervation and reduces the period of denervation, consequently ameliorating denervation-induced muscle atrophy; ii) avoidance of engrafting replacement cells into the neurotoxic environment that exists within the CNS of ALS patients and the necessity for reinnervating axons of CNS-engrafted motor neurons to overcome the inhibitory CNS:PNS barrier in order to exit the CNS and grow the often long distances to target muscles; iii) specificity of optical stimulation to the engrafted ChR2+ motor neurons avoids painful off-target activation of sensory afferents or aberrant activation of endogenous motor axons associated with electrical nerve stimulation (ENS); and critically, iv) the ability to recruit motor units in correct physiological order using optical nerve stimulation (ONS) Llewellyn et al. (2010), avoids the problem of rapid muscle fatigue associated with ENS and incorrect, non-physiological motor unit recruitment. Furthermore, ENS-mediated control of muscle function depends on the presence of surviving motor axons and, since these are progressively lost during the course of disease progression in ALS, the ability of ENS to induce muscle contraction is steadily eroded. More importantly, it has recently been shown that ENS, applied to the phrenic nerve to assist respiratory function in two separate clinical trials in ALS patients McDermott et al. (2016); Gonzalez-Bermejo et al. (2016) accelerated diaphragm muscle denervation, which likely contributed to a significant reduction in patient life-span Guimarães-Costa et al. (2019). Therefore, it is unlikely that ENS will ever be suitable for artificial control of critical muscle function in ALS patients. In contrast, optogenetic stimulation Henderson et al. (2009), in combination with neural replacement, represents a safe alternative approach to artificially restore innervation and function of paralyzed muscles in ALS.

In the current study, we sought to optimize critical elements of this novel therapeutic strategy and to determine whether ChR2+ motor neurons can be used to successfully restore innervation, induce muscle contraction and prevent atrophy of targeted muscles in a model of ALS using the highly aggressive SOD1G93A mouse model of ALS.

Results

Host-vs-graft rejection causes loss of most intraneural ESC-MNs allografts

ChR2+ motor neurons were derived from our previously characterized mouse embryonic stem cell (mESC) line Bryson et al. (2014) and differentiated using a well-established protocol Wichterle et al. (2002). Since these donor ChR2+ motor neurons were generated from an mESC line originating from the 129S1/SvImJ mouse strain (Supplementary Table 1) and recipient mice SOD1G93A mice were on a congenic C57BL/6J genetic background, they constitute an allogeneic source of donor cells. Importantly, allogeneic cells are likely to provide a more cost-effective, off-the-shelf cell therapy platform, compared to generating individual, patient-specific batches of cells suitable for autologous engraftment. Therefore, an important initial objective of this study was to identify an immunosuppression regimen capable of preventing host-vs-graft rejection of allogeneic ChR2+ motor neurons to enable their innervation of target muscles in recipient SOD1G93A mice.

In the absence of any immunosuppression, the survival rate of ChR2+ motor neuron grafts at 35 d post engraftment, was extremely low (<5%) in SOD1G93A mice. However, in the rare cases where the engrafted ChR2+ motor neurons did survive, we observed robust intramuscular axon growth and partial reinnervation of NMJs, even at very late-stage disease (Figure 1A-C and Video 1). Furthermore, acute optical stimulation of the engrafted ChR2+ motor neurons was able to induce tetanic contraction of the target muscle in these SOD1G93A mice (Figure 1D), although these were weak in magnitude. Interestingly, assessment of shorter timepoints revealed that engrafted ChR2+ motor neurons were capable of surviving for up to 14 days (data not shown). It is therefore likely that the poor long-term survival was due to host-vs-graft rejection of the engrafted cells, rather than disease-related toxicity. These findings suggest that if the engrafted motor neurons can evade the host immune response, they can functionally reinnervate target muscles for a therapeutically relevant timescale in this highly aggressive model of ALS.

Engrafted allogeneic ChR2+ motor neuron can survive and robustly reinnervate target muscles in SOD1G93A mice but occurs rarely. (A) Confocal tile-scan of a longitudinal section of the triceps surae muscle from a 135d SOD1G93A mouse, 45d post-engraftment, showing a rare example of graft survival, in the absence of immunosuppression, with robust intramuscular axon projection. (B) Maximum intensity projection (MIP) images of a confocal z-stack through region of interest (dashed box in (A)) and (C), showing NMJs fully (asterisks) and partially (arrow) innervated by ChR2+ motor neuron axons, as well as innervated by endogenous (GFP/YFP-negative, ChAT-positive) motor axons (arrowhead) and fully denervated endplates (circle). (D) Representative recording (n = 1/3 positive responders) showing characteristically weak contractile responses to repetitive 20 Hz optical stimulation.

Tacrolimus (FK506) overcomes ESC-MN allograft rejection but inhibits muscle innervation

In an effort to enhance graft survival, we first tested the calcineurin inhibitor (CNI), tacrolimus (FK506), which is not only routinely used to facilitate solid organ allograft survival in humans but has also been reported to promote axon regeneration after nerve injury in rats Gold et al. (1995). Cohorts of wild-type and SOD1G93A mice (Supplementary Table 2) were treated with FK506 (5mg, kg-1, d-1; dose selected based on evidence of axon regeneration studies) from the time of intraneural ChR2+ motor neuron engraftment; graft survival and muscle reinnervation were then assessed 30-35d post-engraftment. Although FK506 did enable robust ChR2+ motor neuron allograft survival, at the graft site, in all animals examined (Supplementary Table 2), we identified three major problems with this immunosuppressant agent. Firstly, although the engrafted motor neurons were able to survive and project axons along peripheral nerve branches and within the targeted lower hindlimb muscles (Figure 2A), FK506 completely prevented muscle fibre reinnervation, evidenced by lack of response to acute ONS (data not shown) and complete absence of ChR2+/YFP+ NMJs (Figure 2B) in both SOD1G93A (n = 8) and nerve-ligated wild-type mice (n = 4). Secondly, exuberant growth of carry-over pluripotent stem cells led to intraneural tumour formation in most mice (Figure 2C and Supplementary Table 2), which resulted in overt hindlimb motor deficits (Figure 2D and Video 2). Thirdly, in partial agreement with previous reports in rats Aydin et al. (2004), FK506 prevented body mass increase and/or induced body mass decline in a subset (44.4%; n = 8/18) of SOD1G93A mice (Figure 2E and Figure 2-figure supplement 1A); this effect was less evident in wild-type mice (Figure 2-figure supplement 1B). Onset of body mass decline in B6.SOD1G93A mice is highly consistent and typically occurs at 115 days in Hayworth and Gonzalez-Lima (2009), indicating that, at this relatively high dose, FK506 may be preferentially toxic or may exacerbate disease phenotype in SOD1G93A mice.

FK506 enables long term survival of engrafted ChR2+ motor neurons but inhibits muscle reinnervation and exacerbates disease progression in SOD1G93A mice. (A) Representative confocal image showing that GFP/YFP+ axons are able to project within intramuscular branches, following intraneural engraftment of ChR2+ motor neurons and daily immunosuppression with FK506; obtained from a 112d SOD1G93A mouse at 27d post-engraftment. (B) GFP/YFP+ axons fail to reinnervate NMJs despite the proximity of ChR2+ motor axon terminals to denervated endplates. (C) FK506-mediated immunosuppression permits exuberant growth of carry-over pluripotent stem cells that form intraneural tumours (**) within engrafted sciatic nerve (SN) branches; (D) these tumours cause severe movement impairment of the affected hindlimb. (E) FK506 caused body mass loss in a subset (8/18) of SOD1G93A mice, treated at different ages, that precedes onset of normal decline in this model (indicated by vertical dashed line at 115d).

Since peripheral neuropathy is a known adverse event associated with calcineurin inhibitors Arnold et al. (2013), it is possible that FK506 treatment alone may adversely affect endogenous or engrafted motor axons. Indeed, examination of the cross-sectional area distribution of total (ie. sensory and motor) and motor neuron axons in branches of the sciatic nerve in FK506-treated SOD1G93A and WT mice (Figure 3A) revealed a significant loss of axons that affected most axonal calibres in the tibial nerve in wild-type and SOD1G93A mice (Figure 3B and C). A more pronounced loss of total and motor axons, spanning medium to large sized axonal calibres, was observed in the common peroneal nerve branch (Figure 3D and E), which indicates that FK506 can not only exacerbate ongoing motor axon loss in SOD1G93A mice but can also induce motor axon loss even in wild-type mice. Importantly, these neuropathy-like effects appear to be specific to FK-506, since an alternative immunosuppressant, H57-597 mAb (discussed in detail below), did not significantly alter total or motor axon size distribution or total axon counts (Figure 3-figure supplement 1), compared to untreated SOD1G93A mice.

FK506 causes peripheral nerve axonopathy in SOD1G93A and wild-type mice. (A) Representative examples of wild-type (top) and SOD1G93A (bottom) sciatic nerve transverse sections, showing common peroneal nerve (CPN) and tibial nerve (TN), labelled for total axons (βIII tubulin; green) and motor axons (choline acetyl transferase; ChAT; red); automated axon size distribution analysis of (B) total and (C) motor axon in the tibial nerve (TN); (D) total and (E) motor axons in the common peroneal nerve (CPN) in both wild-type and SOD1G93A mice. Data shown as mean; error bars = SEM; 2-way ANOVA analysis: *denotes p ≤0.05; ** denotes p ≤0.0002; *** denotes p ≤0.002; **** denotes p ≤0.00002; significance bars displayed in red highlight changes directly attributable to FK506, independent of genotype.

Since FK506 is known to suppress myoblast proliferation and differentiation Hong et al. (2002) and can cause rare cases of myopathy in humans Breil and Chariot (1999), it is also possible that the FK506-dependent prevention of muscle reinnervation by engrafted ChR2+ motor neurons is due to a muscle specific effect. In any case, these findings clearly show that FK506 is unsuitable as an immunosuppressant to support ChR2+ motor neuron allograft survival and, indeed, suggest that long-term administration at this relatively high dose should be avoided in ALS patients. Importantly, however, the complete protection of engrafted ChR2+ motor neurons by FK506 did confirm that host-vs-graft rejection was responsible for the poor long-term graft survival observed in the absence of immunosuppression, rather than disease related neurotoxicity.

T-cell modulatory immunosuppression confers graft survival and target muscle innervation

In light of the deleterious effects of FK506, and given our aim of conferring compatibility of allogeneic ChR2+ motor neurons as a universal cell therapy for ALS, we sought to identify a more specific form of immunosuppression that avoids the negative effects of FK506 yet supports long term graft survival. Therefore, we investigated the T-cell receptor-beta (TC-beta) targeting monoclonal antibody, mAb H57-597, which has previously been shown to effectively promote long-term heart allograft survival in mice Miyahara et al. (2012) In addition, since our findings indicated that immunosuppression results in a greater risk of tumour formation from carry-over pluripotent stem cells, differentiated ChR2+ motor neurons were also pre-treated with mitomycin-C (MMC; 2 μg/ml for 2 hrs) prior to engraftment, to eliminate tumorigenic cells Magown et al. (2016), to further enhance the translational potential. MMC-treated ChR2+ motor neurons were unilaterally engrafted into the tibial nerve of symptomatic SOD1G93A mice (aged 95.7 ±4.6 days) in conjunction with transient H57-597 mAb delivery (1mg, kg-1; i.p. injection at 0, 1-, 3-, 7- and 14-days post-engraftment). The extent of reinnervation and the ability to optically control the function of the triceps surae (TS) muscle group in the lower hindlimb of SOD1G93A mice was then assessed at late-stage disease (133 ±6.9 d; n = 12) by determining the physiological response of the reinnervated muscles to acute ONS of the engrafted motor neurons followed by histological analysis of muscle and nerve. Histological analysis confirmed that in mAb H57-597 treated animals, engrafted ChR2+ motor neurons were present in all recipient SOD1G93A mice (n > 84 to date). Importantly, there was significant axonal projection within intramuscular nerve branches and robust reinnervation of muscle fibres in the targeted TS muscle (Figure 4A and 4B; Video 3 and 4). As we previously reported in a wild-type mouse nerve ligation model Bryson et al. (2014), some de novo NMJs in SOD1G93A mice exhibited signs of immaturity, including polyinnervation (Figure 4C), as well as collateral and terminal sprouting of motor axons (Figure 4D). Since the peripherally-engrafted reinnervating motor neurons are inactive during the post-engraftment period and progressive muscle atrophy is ongoing, only 10.5% of endplates are innervated by engrafted ChR2+ motor neurons (Figure 4E). Importantly, these reinnervated endplates are functional, since acute in vivo ONS of the engrafted ChR2+ motor neurons in the exposed sciatic nerve of late-stage disease SOD1G93A mice at 133 ±7.2 days of age (37.7 ±5.1 days post-engraftment; n = 12), induced positive contractile responses in all animals, although the amplitude of the maximal contractile force was still weak (0.8 ± 0.2 g; n = 11).

Transient H57-597 mAb treatment confers complete ChR2+ motor neuron allograft survival and, importantly, allows robust triceps surae muscle reinnervation up until late-stage disease in SOD1G93A mice. (A) 3-D reconstruction of 67 individual tile-scans, acquired from serial sections from an entire triceps surae muscle, from a 135d SOD1G93A mouse following engraftment of ChR2+ motor neurons at 95d and H57-597 treatment, showing the full extent of axonal projection throughout the whole muscle; see also Video 1. (B) A high-resolution maximum intensity projection (MIP) image of a confocal z-stack, revealing multiple NMJs innervated to varying extents (α-bungarotoxin; α-BTx; red)) by YFP+ engrafted motor neuron axons (green) labelled for choline-acetyl transferase (ChAT; blue – note; overexposure of blue channel enables visualization of muscle fibres). (C) MIP image of a confocal z-stack showing an example of a partially innervated NMJ. (D) MIP image of a confocal z-stack showing an example of a fully innervated NMJ; note, poly-innervation, shown in (C), and terminal sprouting, shown in (D) which are signs of immaturity. (E) Automated quantification of total endplate number (count = 3,482; labelled with α-BTx) and manual counts of endplates (count = 364) that showed YFP colocalization, indicating innervation, from the same muscle.

Motor neuron subtype identity does not affect response to optical stimulation

In an effort to increase the amplitude of the contractile response of the target muscle to optical stimulation we next tested whether engraftment of motor neurons with a fast-firing subtype identity may be more suitable than engraftment of predominantly slow-firing motor neurons by using alternative differentiation protocols.32 The stronger, predominantly fast-twitch, gastrocnemius component of the TS muscle group is usually innervated by fast-firing/fast-fatigable (FF) motor neurons, which are known to have the capacity to innervate many more individual muscle fibres per motor unit than slow-firing motor neurons, which normally innervate a much smaller number of weaker, slow-twitch muscle fibres, predominantly within the soleus and plantaris regions of the triceps surae. MMC-treated ChR2+ motor neurons, differentiated to yield FF subtype identity motor neurons, were engrafted into the tibial nerve of (106 ±7.2 days) SOD1G93A mice, in combination with transient H57-597 mAb treatment. Maximum isometric muscle contraction of the TS muscle in response to acute optical stimuli was then determined at the same age (133.9 ±7.2 days, n = as previous grafts of predominantly slow-firing motor neurons (133 ±6.9 days, n = 11). This physiological analysis revealed that the motor neuron subtype identity did not significantly affect amplitude of the muscle response to acute optical stimulation (Figure 4-figure supplement 1A) and that the maximum contractile force elicited by ONS remained extremely weak compared to supra-maximal ENS (Figure 4-figure supplement 1B). This result implies that, unlike during normal neuromuscular development, motor neuron subtype identity is not an important determinant of muscle fibre innervation in the mixed fibre type triceps surae muscle. This finding has significance for future clinical translation, since only a single subtype of motor neuron may be required to innervate a variety of different muscles. Only MMC-treated motor neurons with a slow-firing medial motor column identity were used in the subsequent experiments reported here. Since modification of motor neuron subtype identity did not enhance contractile response, our next challenge was to identify an effective method to enhance reinnervation and force generating capacity of the targeted muscle in response to optical stimulation.

Optical stimulation training significantly enhances target muscle force generation

Spinal motor circuit development Milner and Landmesser (1999); Li et al. (2005) and NMJ formation/maintenance Sanes and Lichtman (1999) are known to be activity-dependent processes, thus, without regular stimulation, although peripherally engrafted ChR2+ motor neurons may survive, they are unlikely to form mature NMJs and will therefore have little effect on declining muscle function and atrophy in SOD1G93A mice. Moreover, paralysis and atrophy of affected muscles proceed unchecked in the SOD1G93A mouse model of ALS. We therefore investigated whether regular activation of the engrafted ChR2+ motor neurons, in conjunction with H57-597 mAb treatment, could enhance NMJ maturation and maximize the ability of target muscles to generate contractile force in response to acute optical stimulation in late-stage SOD1G93A mice. To do this, we adapted a wireless, fully implantable optical stimulation system Montgomery et al. (2015), in order to implement a daily optical stimulation training regimen for engrafted mice. First, we modified the printed circuit board (PCB) design to simplify assembly (Figure 5-figure supplement 1A, implemented a new encapsulation method to ensure long term survival of the devices after implantation (Figure 5-figure supplement 1B) and incorporated an RF signal switch and pulse controller to deliver precisely timed RF pulses to power a 470nm light emitting diode (LED) connected to the implantable device (Figure 5-figure supplement 1C).

The modified optical stimulation devices were then surgically implanted in SOD1G93A mice concomitantly with intraneural engraftment of ChR2+ motor neurons, with the trailing LED positioned in close apposition to the graft site (Figure 5A). Commencing at 14 d post-engraftment, to allow growing ChR2+ motor axons sufficient time to reach the target muscle, the mice were transferred to a custom built chamber located above a resonance-frequency induction cavity for 1 hr per day, in order to undertake optical stimulation training (OST; Figure 5B and Video 5), using a bespoke pulse pattern that was empirically determined to elicit maximum contraction (Figure 5-figure supplements 2, 3 and 4), followed by a 2 s rest interval. Following daily OST in engrafted SOD1G93A mice for 21 days, isometric muscle tension physiology was performed at late-stage disease to determine the maximal contractile force of the TS muscle elicited by acute ONS of the exposed sciatic nerve graft site. In confirmation of our hypothesis, there was a highly significant, 9.4-fold, increase in the maximal tetanic force (7.5 ±0.94 g versus 0.8 ±0.2 g; p = ≤0.000001) elicited by ONS in the engrafted OST group of SOD1G93A mice at late-stage disease, compared to age-matched untrained SOD1G93A mice (132.4 ±6.8 days versus 133 ±6.9 days; n = 7 and 11, respectively; Figure 5C-G), although OST did not alter contractile rate characteristics (Figure 5-figure supplement 5A-C). Moreover, quantitative comparison of the maximum force elicited by ONS compared to ENS of TS muscles in late-stage SOD1G93A mice (Figure 5H), showed that in mice that underwent OST, acute ONS elicits up to 22.7% (± 4.5) of total residual muscle force produced by supra-maximal ENS, which activates both endogenous and engrafted motor neurons (Figure 5I), in contrast to only 1.46% (± 0.18) in untrained SOD1G93A mice (Figure 5-figure supplement 6A and 6B). This represents a remarkable >13-fold improvement in force generation. In engrafted SOD1G93A mice that did not undergo OST, the extremely weak twitch contractions in response to ONS precluded interrogation of individual motor unit force values and determination of motor units number estimates (MUNE) in most mice. In contrast, in SOD1G93A mice that underwent OST, the significantly increased contractile response to ONS enabled clear separation of individual motor unit values (Figure 5J), enabling MUNE values to be determined (Figure 5-figure supplement 5D). Furthermore, as we previously reported in nerve-ligated WT mice,22 delivery of repetitive ONS pulses (250 ms bursts of 20 Hz illumination, every 1 s, for 180 s duration) to induce rhythmic, submaximal contraction of the TS muscle did not induce rapid muscle fatigue, whereas equivalent pulses of ENS stimulation of the contralateral TS muscle did result in rapid muscle fatigue (Figure 5K). This observation has significant implications for the ability to control repetitive motor functions in ALS patients.

Daily optical stimulation training (OST) of post-symptom onset SOD1G93A mice engrafted with ChR2+ motor neurons, significantly enhances contractile response to optical stimulation. (A) Schematic indicating intraneural engraftment site in distal tibial nerve and reinnervated triceps surae (TS) muscle, along with stimulation device (top inset) implantation site and subcutaneous LED position; +4 div MMC-treated ChR2+ motor neurons express YFP (green; lower inset box); experimental timescale is shown below. (B) Still frame (taken from Video 5) showing daily OST. Representative isometric muscle tension physiology recordings from the TS muscle in response to specified pulses of ONS in untrained (C) and +OST (D) late-stage SOD1G93A mice, along with ENS response from the same muscle (E). (F) Delivery of an optimized pulse pattern elicits maximal response to ONS that can be used to finely control repetitive contractions; (G) Dashed box is shown at higher temporal resolution to indicate square-wave TTL pulse pattern (that drives LED stimulator) and an individual tetanic contraction. (H) Quantification of maximum contractile responses to indicated pulse patterns of acute ONS shows a highly significant improvement in force generation in late-stage SOD1G93A mice that underwent OST versus untrained controls (dashed horizontal line indicates maximum value from our previous study in nerve ligated WT mice). (I) The proportion of total muscle capacity (determined by supramaximal ENS minus ONS value) elicited by acute ONS is also significantly higher following OST; data represent mean ± SEM. (J) Motor unit number estimate (MUNE) traces obtained from a representative late-stage SOD1G93A mouse, following OST, in response to ipsilateral ONS and ENS, along with contralateral ENS (note, different scales). (K) Fatigue trace recordings comparing ipsilateral ONS (top) and contralateral ENS (bottom) in the same late-stage SOD1G93A mouse, in response to 250ms 20Hz pulses, repeated every 1s for 180s.

Optical stimulation training prevents atrophy of reinnervated muscle fibres

Finally, having established that optical stimulation training significantly enhances the maximal force elicited by optical stimulation of engrafted ChR2+ motor neurons in late-stage SOD1G93A mice, we examined whether long-term optical stimulation training could also prevent atrophy of reinnervated muscle fibres. Since NMJs comprise an extremely small volume of the entire muscle, high-resolution 3D imaging of the entire muscle to determine muscle fibre innervation status and fibre diameter information is not feasible. Therefore, we developed a novel technique, termed digital Cross-sectional area Analysis from Longitudinal Muscle Sections (dCALMS), in order to assess these properties. This involved 3D reconstruction and analysis of regions of interest, obtained from 30μm thick longitudinal TS muscle sections (Figure 6A) from ChR2+ motor neuron engrafted, late-stage SOD1G93A mice that had undergone OST. Each region contained at least one NMJ innervated by a ChR2+ motor neuron, along with randomly captured neighbouring fibres (Figure 6B, Figure 6-figure supplement 1 and Video 6). The 3D reconstructions were then digitally re-sliced in the transverse orientation (Figure 6C, top panel), in order to obtain data on muscle fibre cross-sectional area (CSA) using a semi-automated process (Figure 6C, lower panel). The dCALMS analysis revealed that the average CSA of muscle fibres innervated by engrafted ChR2+ motor neurons were similar in size to fibres still innervated by residual endogenous motor neurons (922.3 versus 1018.3 μm2; p = ≤0.85; Figure 6D). Importantly, the CSA of muscle fibres innervated by ChR2+ motor neurons was significantly greater than fibres with completely denervated endplates (average CSA = 525.4 μm2; p = ≤0.00001) or fibres whose innervation status could not be determined (average CSA = 668.8 μm2; p ≤0.0003), since the endplate was outside the scanned region of interest.

Optical stimulation training prevents atrophy of muscle fibres that have been reinnervated by ChR2+ motor neurons in late-stage SOD1G93A mice. (A) Confocal tile-scan showing a single longitudinal section through the triceps surae muscle of 135d SOD1G93A mouse (35d post-engraftment), following daily OST; endplates (labelled with αBTx) innervated by GFP/YFP+ engrafted motor neurons are evident throughout the whole muscle group, including fast-twitch medial gastrocnemius (MG) and lateral gastrocnemius (LG) muscles and the slow-twitch soleus (Sol) muscle. (B) Representative top-down maximum intensity projection (MIP) view of a confocal z-stack through a 30μm longitudinal section obtained from the same mouse; including side-on (B’) and end-to-end (B”) MIP views of the same z-stack; a de novo NMJ, innervated by a ChR2+ motor neuron (asterisk), is indicated on a muscle fibre that also has a denervated endplate (arrow), along with another muscle fibre still innervated by an endogenous choline-acetyltransferase (ChAT+) positive, GFP-negative motor neuron (arrowhead). (C) A digital slice (top) through the 3-D z-stack obtained at the y-axis plane indicated by dashed line in (B) and colorized masks delineating individual muscle fibres (bottom); see Video 6. (D) Average cross-sectional area of individual muscle fibres with an innervation status that was unconfirmed (117 fibres), innervated by GFP+ motor neurons (62 fibres), denervated (28 fibres), or innervated by endogenous motor axons (13 fibres); n = 3 late-stage engrafted SOD1G93A mice that had undergone OST; Data shown as mean; error bars = SEM; one-way ANOVA with Tukey’s post hoc correction: *denotes p ≤0.05; ** denotes p ≤0.0002; *** denotes p ≤0.002; **** denotes p ≤0.00002.

Discussion

This study shows for the first time that replacement stem cell-derived motor neurons can robustly and reliably reinnervate target muscles in the highly aggressive SOD1G93A mouse model of ALS, even when engrafted after onset of overt symptoms; moreover, the restored innervation can be maintained even until extremely late-stage disease. Furthermore, our findings suggest that engrafted ChR2+ motor neurons can not only provide an interface to safely and selectively control the function of targeted muscle but, also, that regular optical stimulation training (OST) can be used to: i) reinforce connectivity between engrafted motor neurons and muscle fibres; ii) significantly enhance the maximal force elicited by optical stimulation of the targeted muscle; and iii) prevent atrophy of muscle fibres that are reinnervated by engrafted motor neurons. The highly significant improvements in muscle innervation, atrophy prevention and maximum contractile force, as a result of the daily OST regimen, confirms that stimulation-induced activity is necessary to maximize connectivity between engrafted motor neurons and their target muscles.

The prevailing view in the ALS field is that spinal motor neuron pathology first manifests at the nerve terminal and that muscle fibres are likely to actively contribute to degeneration of endogenous motor neurons in ALS Scaricamazza et al. (2021). Our findings clearly demonstrate that affected muscles, even in the highly aggressive SOD1G93A ALS model, remain receptive to reinnervation by healthy engrafted motor neurons, even until late-stage disease. Moreover, once target muscles have been reinnervated, this approach enables implementation of a safe muscle training/exercise regimen that can be used to preserve muscle integrity and prevent irreversible muscle wasting that otherwise occurs as a result of progressive neurodegeneration in ALS Mora (1989). Since skeletal muscles are not simply biomechanical actuators, but have complex functions in overall metabolic homeostasis, thermoregulation, venous return and maintenance of blood volume, the ability to prevent muscle atrophy using OST, is likely, by itself, to have major health benefits for ALS patients. In contrast, the use of electrical nerve stimulation (ENS) to control muscle function or, indeed, ENS-based exercise programs is likely to accelerate degeneration of remaining motor axon terminals Guimarães-Costa et al. (2019), and is therefore unlikely to be safe. Unlike ENS, the highly selective nature of ONS does not activate or interfere with endogenous motor neuron function and due to its ability to recruit motor units in physiological order, ONS has the added significant benefit of avoiding rapid muscle fatigue Llewellyn et al. (2010). The ability to combat muscle atrophy, using OST, could extend the ability of targeted to execute functionally useful movements, potentially indefinitely in ALS patients.

Since differentiation methods that yield either fast-firing or slow-firing motor neurons did not appear to affect the ability of engrafted motor neurons to innervate the mixed fibre type TS muscle group in SOD1G93A mice, motor neuron subtype identity appears to be redundant in this case. Therefore, a single type of motor neuron, produced at scale, could potentially be used to target a large number of different muscles in each patient. This has advantages in terms of simplifying the regulatory approval process, since the donor motor neurons would be produced in exactly the same way, irrespective of the graft site or recipient.

Indeed, another key finding of this study that could streamline future translation of this therapeutic strategy, is the identification of a TCR-β targeting antibody, H57-597 mAb, as an effective mediator of allograft survival. Existing T-cell targeting monoclonal antibodies, such as OKT-3, have been clinically approved Page et al. (2013) and, importantly, this form of immunosuppression overcomes severe adverse effects that we observed with the commonly used CNI-based immunosuppressant, tacrolimus (FK506). Our data shows that H57-597 mAb treatment was well tolerated during transient administration from symptom-onset up until late-stage disease in SOD1G93A mice, however, the aggressive disease progression in this model precludes investigation of longer-term effects and it remains to be seen how long-term TCR-β-based immunosuppression may be tolerated in ALS patients. A similar immunosuppression regimen involving a CD25 (IL2) targeting monoclonal antibody, Basiliximab, along with glucocorticoid and tacrolimus maintenance therapy, has been shown to be safe but poorly tolerated in ALS patients, as an independent investigative approach Fournier et al. (2018), as well as part of a separate clinical trial assessing intraspinal grafts of neural precursor cells in ALS patients Mazzini et al. (2019). Therefore, immunosuppression per se, is not an impediment to this strategy.

The use of allogeneic donor cells, with a safe and effective immunosuppression regimen, means that a future cell therapy could potentially be universally compatible with all ALS patients, which would significantly reduce costs and simplify the regulatory approval process, compared to individually tailored autologous cell grafts. Of course, further studies will be required to ensure that human-compatible, induced pluripotent stem cell (iPSC)-derived donor motor neurons are able to function in the same manner as allogeneic mouse ESC-derived motor neurons. The generation of HLA-matched super donor hiPSC lines may further mitigate the need for immunosuppression Turner et al. (2013), however, immunogenicity of the ChR2 protein could mean that some form of immunosuppression may be necessary for any optogenetic therapy in the peripheral nervous system, including viral delivery Maimon et al. (2018).

The anatomical separation of specific nerve branches in humans means that this approach could be used to target and independently control large numbers of muscles in each individual ALS patient. However, it will be first necessary to demonstrate safety and efficacy in a relatively low risk muscle to restore a simple motor function. For example, the common peroneal nerve is highly accessible from a neurosurgical perspective and existing ENS devices, developed to correct foot drop for other neurological disorders Hausmann et al. (2015), could be readily adapted to assist ambulatory function in early-stage ALS patients. In the longer term, existing multichannel ENS devices, which have been developed to control more complex ADLs in high-level spinal cord injury (SCI) patients Memberg et al. (2014), could also be adapted into a minimally invasive, transcutaneous optical stimulation device Maimon et al. (2017). This combinatorial therapeutic strategy, comprising allogeneic donor cells, an effective immunosuppression regimen and optical stimulation device, is highly compatible with the rapidly evolving field of brain computer interface (BCI) technology. BCI could be used to decode a paralyzed patient’s intention to perform a given movement in order to control the activity of the engrafted motor neurons, which provide the necessary interface to execute the intended movement via a wearable optical stimulation device. This novel approach would entirely bypass the severe damage that occurs throughout the entire CNS in ALS patients and enable autologous control of movement. Moreover, this strategy also has broad utility for a wide range of other neurogenic causes of paralysis, such as spinal cord injury and stroke.

Although the findings of this study clearly demonstrate that our combinatorial cell therapy approach is effective in a highly aggressive mouse model of ALS, further investigation is required in order to confirm that the strategy can be applied to alternate model of ALS, with a longer lifespan, in order to fully explore the long-term efficacy of the approach, particularly in terms of chronic allograft survival using a transient immunosuppression approach. It is possible that some form of maintenance therapy may also be required to confer long term graft survival. Of course, the biggest challenge will be to demonstrate that human optogenetically-modified motor neurons, derived from either induced pluripotent stem cells (iPSCs) or human ESCs, are capable of reinnervating target muscles in the same manner as we have demonstrated for mESC-derived motor neurons. It will also be necessary to scale this cell therapy strategy up, using larger animal models that more accurately recapitulate human-scale anatomy.

Despite these remaining challenges, the findings of this study provide strong support for this novel cell therapy, which, if successful, could finally begin to deliver major health benefits for ALS patients.

Methods and Materials

Detailed methods are provided in the Supplementary material.

mESC motor neuron differentiation

The Channelrhodopsin2-YFP expressing mESCs (Clone C9G) used in this study were generated as previously describedBryson et al. (2014). Motor neuron differentiation was performed according to a standard protocol developed by Wichterle et al Wichterle et al. (2002) for production of predominantly slow-firing medial motor column (MMC) identity motor neurons and an updated “caudalized-ventralized” (CV) protocol, also developed by the Wichterle group Peljto et al. (2010), that produces a higher proportion of motor neurons with fast-firing properties. Following differentiation on Day 5 (or Day 7 for the C-V protocol), embryoid bodies (EBs) containing differentiated motor neurons were dissociated and total cells were resuspended in PBS at a concentration of 50,000 cell/μl. Where indicated, mitomycin-C (1μg/μl) was added to the EBs for 2 hrs prior to dissociation. Nile Blue A (0.0002% final concentration) was added to the cell suspension and the cells were kept on ice until engraftment. Single nucleotide polymorphism (SNP) analysis of mESC lines (Clone C9G, HBG3 and a C57BL/6J mESC line for reference) was carried out by Charles River Laboratories. See the Supplementary materials for full details 1.

Mice

All procedures and experiments involving animals were carried out under License from the UK Home Office in accordance with the Animals (Scientific Procedures) Act 1986 (Amended Regulations 2012), following ethical approval from the UCL Queen Square Institute of Neurology Animal Welfare Ethical Review Body (AWERB), and in accordance with the ARRIVE guidelines. B6.CgTg(SOD1*G93A)1Gur/J mice (The Jackson Laboratory, stock number 004435) were bred specifically for this study by mating presymptomatic male transgene carriers with congenic C57BL/6J females (Charles River Laboratories)1.

Intraneural engraftment of mESC-derived ChR2+ motor neurons

Surgical engraftment of ChR2+ motor neurons was performed under aseptic conditions, as previously described Bryson et al. (2014). Briefly, 1μl of dissociated EB cell suspension (50,000 cells) was injected into the tibial nerve close to the trifurcation point of the sciatic nerve, using a 5μl Hamilton syringe equipped with a customized 33G needle. Where indicated, an implantable optical stimulation device was inserted through the same surgical incision and positioned subcutaneously under the skin on the back; the trailing LED was fixed with sutures to the overlying muscles at the graft site during wound closure. Immunosuppression, as indicated, was initiated at the time of surgical engraftment. See supplementary materials for full details1.

Implantable optical stimulation devices and power transmission system

Optical stimulation devices were largely produced as described by Montgomery et al. Montgomery et al. (2015), with minor, but essential, modifications to the PCB design and encapsulation method. Similarly, the power transmission system used to activate the implanted LED devices was also largely as described by Montgomery et al. Montgomery et al. (2015), however, a Solid State Switch (MiniCircuits; ZX80-DR230-S+), controlled by a USB-TTL Interface (Prizmatix), was used to generate specific optical stimulation patterns. Engrafted SOD1G93A mice that underwent daily optical stimulation training were placed in the stimulation chamber for 1 hr/day from post-engraftment day 14 until termination of the experiment at late-stage disease (132.4 ±6.8 days). See the Supplementary materials for full details1.

Isometric muscle tension physiological analysis

At the experimental end-point, engrafted SOD1G93A mice underwent isometric muscle tension physiology, in order to accurately determine the contractile properties of the triceps surae muscle in response to acute optical stimulation, as previously described Bryson et al. (2014), with the following modifications: a PowerLab 4/30 stimulation and recording system (AD Instruments) was used to deliver bespoke electrical stimulation signals, either as direct constant voltage pulses applied to the nerve via bipolar electrodes for electrical nerve stimulation (ENS), or used as a 5V TTL signal to activate a 470nm LED light-source (CoolLED; pe100), delivered to the exposed sciatic nerve via a liquid lightguide for optical nerve stimulation (ONS); stimulation program available on request. LabChart software (AD Instruments) was used for automated data analysis of contractile parameters. See the Supplementary materials for full details1.

Nerve and muscle histology and image analysis

See Supplementary materials for full details, including automated motor/sensory axon CSA analysis1, axon counts and innervation analysis method1. The digital CSA analysis of longitudinal muscle sections (dCALMS) method is also described in full in the Supplementary materials1.

Quantification and statistical analysis

Sample sizes

The number of mice (n) is provided in the figures and/or figure legends; also see Supplementary Tables 2 and 3.

Statistical analysis

All data are presented as mean ± SEM unless otherwise indicated. GraphPad Prism 9 (Prism) was used for statistical analyses. No out-liners or data points were eliminated. Differences between two groups were assessed using multiple two-tailed unpaired t tests. Differences between more than two groups were assessed by using one-way or two-way analysis of variance (ANOVA) with multiple comparison correction, or mixed model effects analysis, as stated in the figure legends. Significance was defined as *P ≤0.05, **P ≤0.01, ***P ≤0.001, or ****P ≤0.0001. See supplementary Methods for further details 1.